Analyzing Performance Bottleneck in Graph Neural Network Training: An Experimental View

### 2020-07-23 Manuscript/sec1-introduction.tex
统一为edge/vertex表达

To find out the performance bottleneck in the GNN training, we conduct experimental analysis in this work.
We model the typical GNNs with the message-passing framework.
Since we focus on the efficiency issue of GNN training instead of accuracy, we classify the typical GNNs into four quadrants according to their vertex and edge calculation time complexity.
We pick GCN \cite{kipf2017_gcn}, GGNN \cite{li2015_ggnn}, GAT \cite{huang2018_gat} and GaAN \cite{zhang2018_gaan} as representative GNNs with different vertex/edge calculation complexities.
We implement them with PyG and evaluate their efficiency with six real-world datasets on GPU.
We verify the correctness of the complexity analysis by measuring the effects of the hyper-parameters on the training time and memory usage.
We find out the most time-consuming stage in GNN training by decomposing the training time per epoch from the layer level to the operator level.
We analyze the memory usage during the training to discover the main factor that limits the data scalability of GNN training on GPU.
Finally, we evaluate whether the sampling techniques affect the performance bottleneck.
Our key findings and insights are summarized below.

>>>>

To find out the performance bottleneck in the GNN training, we conduct experimental analysis in this work.
We model the typical GNNs with the message-passing framework.
Since we focus on the efficiency issue of GNN training instead of accuracy, we classify the typical GNNs into four quadrants according to their vertex and edge calculation time complexity.
We pick GCN \cite{kipf2017_gcn}, GGNN \cite{li2015_ggnn}, GAT \cite{huang2018_gat} and GaAN \cite{zhang2018_gaan} as representative GNNs with different edge/vertex calculation complexities.
We implement them with PyG and evaluate their efficiency with six real-world datasets on GPU.
We verify the correctness of the complexity analysis by measuring the effects of the hyper-parameters on the training time and memory usage.
We find out the most time-consuming stage in GNN training by decomposing the training time per epoch from the layer level to the operator level.
We analyze the memory usage during the training to discover the main factor that limits the data scalability of GNN training on GPU.
Finally, we evaluate whether the sampling techniques affect the performance bottleneck.
Our key findings and insights are summarized below.

### 2020-07-23 Manuscript/sec4-experiments.tex
统一为edge/vertex表达

    \caption{Effects of hyper-parameters on the vertex/edge calculation time.}
    \label{fig:exp_hyperparameter_on_vertex_edge_phase_time}
\end{figure}

@@

\begin{figure}
    \centering
    \subfloat[GCN]{\includegraphics[height=4cm]{figs/experiments/exp_avg_degree_on_vertex_edge_cal_time_gcn.pdf}}
    \subfloat[GGNN]{\includegraphics[height=4cm]{figs/experiments/exp_avg_degree_on_vertex_edge_cal_time_ggnn.pdf}}\\
    \subfloat[GAT]{\includegraphics[height=4cm]{figs/experiments/exp_avg_degree_on_vertex_edge_cal_time_gat.pdf}}
    \subfloat[GaAN]{\includegraphics[height=4cm]{figs/experiments/exp_avg_degree_on_vertex_edge_cal_time_gaan.pdf}}
    \caption{Effects of the average degree on the time proportion of the vertex/edge calculation. Graphs were generated with the R-MAT generator by fixing the number of vertices as 50,000. }
    \label{fig:exp_avg_degree_on_vertex_edge_cal_time}
\end{figure}

The experimental results also indicate that the average degree of the dataset affects the time-consuming proportion of the vertex/edge calculation.
For GaAN, the time spent on the vertex calculation exceeded the edge calculation on the \texttt{pub} and \texttt{cam} datasets, because the average degrees of the two datasets were low, making $|\mathcal{E}|$ and $|\mathcal{V}|$ much closer.
To evaluate the effects of the average degree, we used the R-MAT model to generate random graphs with 50k vertices and the average degrees ranging from 2 to 100.
\figurename~\ref{fig:exp_avg_degree_on_vertex_edge_cal_time} shows the training time of the four GNNs under different average degrees.
As the average degree increased, the training time of the edge calculation grew \emph{linearly}.
For GCN, GAT, and GaAN, the edge calculation dominated the entire training time even under small average degrees.
Only for GGNN that had high vertex and low edge calculation complexities, the training time of the vertex calculation could exceed the edge calculation under low average degrees ($<5$).
Therefore, \emph{improving the efficiency of the edge calculation is the key to reduce the GNN training time}.

@@

The average degree of the graph also affects MER by influencing the relative scale of the intermediate results from the vertex/edge calculation.
Fixing the number of vertices $|\mathcal{V}|$, we used the R-MAT generator to generate random graphs with different average degrees.
\figurename~\ref{fig:exp_memory_expansion_ratio_input_graph_number_of_edges} shows how the memory usage changes according to the average degrees.
As the average degree $\bar{d}$ increased, the number of edges $|\mathcal{E}|$ increased and the peak memory usage increased \emph{linearly} with $\bar{d}$.
The edge calculation gradually dominated the memory usage and \emph{the MER converged to a stable value}.
The stable value was determined by the complexity of the edge calculation.
Except for GGNN, the MER of the other GNNs increased as $\bar{d}$ increased.
As GGNN had high vertex calculation complexity, the MER related to the vertex calculation was much larger than the edge calculation.
When the edge calculation dominated the memory usage, its MER became smaller.

>>>>

    \caption{Effects of hyper-parameters on the edge/vertex calculation time.}
    \label{fig:exp_hyperparameter_on_vertex_edge_phase_time}
\end{figure}

@@

\begin{figure}
    \centering
    \subfloat[GCN]{\includegraphics[height=4cm]{figs/experiments/exp_avg_degree_on_vertex_edge_cal_time_gcn.pdf}}
    \subfloat[GGNN]{\includegraphics[height=4cm]{figs/experiments/exp_avg_degree_on_vertex_edge_cal_time_ggnn.pdf}}\\
    \subfloat[GAT]{\includegraphics[height=4cm]{figs/experiments/exp_avg_degree_on_vertex_edge_cal_time_gat.pdf}}
    \subfloat[GaAN]{\includegraphics[height=4cm]{figs/experiments/exp_avg_degree_on_vertex_edge_cal_time_gaan.pdf}}
    \caption{Effects of the average degree on the time proportion of the edge/vertex calculation. Graphs were generated with the R-MAT generator by fixing the number of vertices as 50,000. }
    \label{fig:exp_avg_degree_on_vertex_edge_cal_time}
\end{figure}

The experimental results also indicate that the average degree of the dataset affects the time-consuming proportion of the edge/vertex calculation.
For GaAN, the time spent on the vertex calculation exceeded the edge calculation on the \texttt{pub} and \texttt{cam} datasets, because the average degrees of the two datasets were low, making $|\mathcal{E}|$ and $|\mathcal{V}|$ much closer.
To evaluate the effects of the average degree, we used the R-MAT model to generate random graphs with 50k vertices and the average degrees ranging from 2 to 100.
\figurename~\ref{fig:exp_avg_degree_on_vertex_edge_cal_time} shows the training time of the four GNNs under different average degrees.
As the average degree increased, the training time of the edge calculation grew \emph{linearly}.
For GCN, GAT, and GaAN, the edge calculation dominated the entire training time even under small average degrees.
Only for GGNN that had high vertex and low edge calculation complexities, the training time of the vertex calculation could exceed the edge calculation under low average degrees ($<5$).
Therefore, \emph{improving the efficiency of the edge calculation is the key to reduce the GNN training time}.

@@

The average degree of the graph also affects MER by influencing the relative scale of the intermediate results from the edge/vertex calculation.
Fixing the number of vertices $|\mathcal{V}|$, we used the R-MAT generator to generate random graphs with different average degrees.
\figurename~\ref{fig:exp_memory_expansion_ratio_input_graph_number_of_edges} shows how the memory usage changes according to the average degrees.
As the average degree $\bar{d}$ increased, the number of edges $|\mathcal{E}|$ increased and the peak memory usage increased \emph{linearly} with $\bar{d}$.
The edge calculation gradually dominated the memory usage and \emph{the MER converged to a stable value}.
The stable value was determined by the complexity of the edge calculation.
Except for GGNN, the MER of the other GNNs increased as $\bar{d}$ increased.
As GGNN had high vertex calculation complexity, the MER related to the vertex calculation was much larger than the edge calculation.
When the edge calculation dominated the memory usage, its MER became smaller.
